\chapter{EXPECTED OUTCOME}

The project aims to deliver a working research prototype for short-term PM$_{2.5}$ forecasting in selected Kathmandu Valley stations with explainable spatio-temporal modeling. The proposed system is expected to support air quality forecasting, interpretation, visualization, and decision-support analysis.

\section{Technical Outcomes}

The expected technical outcomes of the project are as follows:

\begin{itemize}
    \item \textbf{Cleaned and Documented Dataset:}  
    A cleaned and well-documented station dataset with station metadata, hourly pollutant observations, meteorological variables, preprocessing records, and model-ready input features.

    \item \textbf{Preprocessing Pipeline:}  
    A structured preprocessing framework for timestamp synchronization, missing value handling, outlier detection, feature scaling, temporal window generation, and integration of OpenAQ and Open-Meteo data.

    \item \textbf{Wind-Aware Dynamic Graph Module:}  
    A dynamic graph construction method that combines geographical distance, PM$_{2.5}$ correlation, wind direction alignment, and wind speed to model changing station-to-station influence.

    \item \textbf{Virtual Node Estimation Mechanism:}  
    A graph-based method for estimating PM$_{2.5}$ concentration at unmonitored map locations using nearby real monitoring stations and wind-aware spatial relationships.

    \item \textbf{Baseline and Primary Forecasting Models:}  
    Baseline models and a primary GAT-GRU forecasting model evaluated using standard regression metrics and peak-event detection metrics.

    \item \textbf{Explainability Module:}  
    An explanation component that provides station influence analysis, SHAP-based feature importance, and temporal sensitivity analysis for interpreting model predictions.

    \item \textbf{Interactive Dashboard:}  
    An interactive dashboard showing PM$_{2.5}$ forecasts, AQI categories, spatial maps, station-wise trends, virtual-node estimates, and explanation visualizations.

    \item \textbf{Complete Project Report:}  
    A complete project report including methodology, implementation details, performance analysis, limitations, feasibility discussion, and future enhancement possibilities.
\end{itemize}
