\chapter{FEASIBILITY ANALYSIS}

\section{Technical Feasibility}
The project is technically feasible because the required data sources, algorithms, and software tools are available. Air quality data can be collected from OpenAQ, while meteorological data such as wind speed, wind direction, temperature, humidity, pressure, and dew point can be obtained from Open-Meteo. These datasets can be integrated at an hourly resolution and used for PM$_{2.5}$ forecasting in Kathmandu Valley.

The proposed model is also technically feasible because it uses established machine learning and deep learning tools. Data preprocessing can be performed using Python, pandas, NumPy, and scikit-learn. The forecasting models can be implemented using PyTorch and PyTorch Geometric. Baseline models such as persistence models, regression-based models, and \gls{gru} can be developed first, followed by the primary GAT-GRU model.

The use of a wind-aware dynamic graph provides sufficient technical depth for the project. The graph will represent monitoring stations as nodes and station relationships as edges. Edge weights will be constructed using geographical distance, PM$_{2.5}$ correlation, wind direction, and wind speed. This allows the model to consider both spatial proximity and meteorological influence when forecasting PM$_{2.5}$ concentration.

The virtual node component also adds research value while remaining technically manageable. It can be validated using a leave-one-station-out strategy, where one real monitoring station is temporarily treated as an unmonitored location and the predicted value is compared with its actual PM$_{2.5}$ reading. This makes virtual-node evaluation possible even when ground truth is not available for every unmonitored location.

\section{Operational Feasibility}
The proposed system is operationally feasible as a research prototype. The system will operate as a web-based dashboard where users can view station-wise PM$_{2.5}$ forecasts, \gls{aqi} categories, spatial pollution maps, and explanation outputs. The backend service can provide processed data, model predictions, and explanation results through REST APIs, while the frontend can display the results through charts, maps, and station-level summaries.

The system does not require physical sensor deployment because it will initially use publicly available and downloadable air quality and meteorological data. This makes the project practical within an academic setting. The dashboard will be designed for interpretation, demonstration, and academic analysis rather than official public warning.

The operational workflow is also practical. Data can be collected and cleaned offline, models can be trained and evaluated using prepared datasets, and the trained model outputs can be displayed through the dashboard. This makes the system suitable for proposal defense, progress demonstration, and final project presentation.

\section{Economic Feasibility}
The project is economically feasible because most of the required tools and datasets are free or open-source. Python, pandas, NumPy, scikit-learn, PyTorch, PyTorch Geometric, FastAPI, React, and SHAP can be used without licensing cost. Public data sources such as OpenAQ and Open-Meteo can be accessed without major financial investment.

The main possible expenses are internet access, optional cloud or \gls{gpu} usage, hosting for dashboard demonstration, printing, and documentation. The computational requirement is expected to be manageable because the model will be developed using selected Kathmandu Valley monitoring stations and hourly observations. Free platforms such as Google Colab, Kaggle Notebook, or local machines can be used during early experiments. The estimated budget is shown in \cref{tab:project_budget}.

\section{Schedule Feasibility}
The project is feasible within the academic project timeline if developed in phases. The first phase will focus on collecting and cleaning PM$_{2.5}$ and meteorological data for selected Kathmandu Valley stations. The second phase will develop baseline models such as persistence, regression-based models, and \gls{gru}. The third phase will implement wind-aware graph construction and the GAT-GRU model. The fourth phase will add virtual-node estimation and explainability. The final phase will integrate the trained model with a dashboard and complete testing, documentation, and presentation.

The focused project scope improves schedule feasibility by allowing the team to concentrate on the most important technical components: data preparation, wind-aware graph construction, PM$_{2.5}$ forecasting, virtual-node estimation, explainability, and dashboard visualization. This phased approach helps ensure that a working prototype can be completed within the academic project timeline while still maintaining sufficient research depth.

\section{Risk and Mitigation}
The major risks of the project are related to data quality, station availability, missing values, weather-data mismatch, model complexity, virtual-node validation, and dashboard integration. These risks and their mitigation strategies are summarized in \cref{tab:risk_mitigation}.

\begin{table}[H]
    \centering
    \small
    \caption{Project Risks and Mitigation Strategies}
    \label{tab:risk_mitigation}
    \renewcommand{\arraystretch}{1.2}
    \setlength{\tabcolsep}{4pt}

    \begin{tabularx}{0.94\linewidth}{|
        >{\raggedright\arraybackslash}p{3.4cm}|
        >{\raggedright\arraybackslash}X|}
        \hline
        \textbf{Risk} & \textbf{Mitigation Strategy} \\
        \hline

        Missing or inconsistent PM$_{2.5}$ readings &
        Select stations with sufficient data availability, apply timestamp alignment, use interpolation, and remove unreliable records. \\
        \hline

        Weather-data mismatch &
        Match each monitoring station with the nearest available weather grid point and align meteorological records by timestamp. \\
        \hline

        Limited station density &
        Start with selected stations that provide useful spatial coverage across Kathmandu Valley and validate graph behavior carefully. \\
        \hline

        Virtual-node validation difficulty &
        Use leave-one-station-out validation before applying virtual-node prediction to completely unmonitored locations. \\
        \hline

        Model complexity &
        Implement simple baseline models first, followed by \gls{gru}, wind-aware graph construction, and then the GAT-GRU model. \\
        \hline

        Long training time &
        Use selected Kathmandu Valley stations, hourly observations, efficient batching, and optional free \gls{gpu} resources only when needed. \\
        \hline

        Dashboard integration complexity &
        Build a simple dashboard with station-wise forecasts first, then gradually add \gls{aqi} categories, maps, explanation plots, and virtual-node outputs. \\
        \hline
    \end{tabularx}
\end{table}

\section{Limitations}
The proposed system has some limitations. Prediction accuracy will depend on the quality, continuity, and availability of station data and meteorological data. Weather variables obtained from external sources may not perfectly represent local conditions at every monitoring station. Virtual-node predictions are estimates and may not fully replace real sensor measurements at unmonitored locations.

The model explanations will provide useful interpretation of station influence, feature importance, and temporal patterns, but they should be understood as model-based explanations rather than absolute physical causation. The system is intended as an academic research prototype for PM$_{2.5}$ forecasting, explanation, and visualization. It will not replace official government air quality monitoring or warning systems.

\section{Final Feasibility Statement}
The proposed project is feasible as a major project because it has a focused and realistic scope while still maintaining strong technical depth. The use of wind-aware dynamic graph construction, GAT-GRU forecasting, virtual-node estimation, and explainability provides a meaningful research contribution for PM$_{2.5}$ forecasting in Kathmandu Valley.

The required datasets, software tools, and development frameworks are available, and the project can be completed in phases within the academic timeline. Therefore, the proposed system is technically, operationally, economically, and schedule-wise feasible as a research prototype for explainable PM$_{2.5}$ forecasting and visualization in Kathmandu Valley.
