\chapter{LITERATURE REVIEW}

\section{Air-Quality Dynamics and Modeling Requirements in Kathmandu Valley}

\subsection{Topography, Monitoring Coverage, and Long-Term Trends}

Kathmandu's topography and sparse monitoring complicate surface-air-quality assessment. Mahapatra et al. reported increasing dry-season aerosol loading during 2000-2015 and substantial background contributions, while chemical-transport simulations underestimated observed valley pollution \cite{ref1}. Thus, even physics-based regional models may remain biased when local terrain, emissions, and boundary conditions are insufficiently represented.

Becker et al. combined ground PM$_{2.5}$, MODIS, CAMS, and MERRA-2 data. Surface observations showed winter maxima, summer minima, morning and evening peaks, whereas satellite products did not resolve the same urban pattern; CAMS was more consistent but still missed important peaks \cite{ref2}. Multi-source data are therefore useful but not interchangeable and require explicit alignment and validation.

Together, \cite{ref1} and \cite{ref2} indicate that ground stations should provide the primary PM$_{2.5}$ supervision, with remote or modeled products used as auxiliary inputs. They also show that an unmonitored Kathmandu location may occupy a distinct topographic, emission, or transport regime rather than representing a simple missing value.

\subsection{Seasonal, Diurnal, and Meteorological Controls}

Kathmandu studies consistently identify nonlinear and lagged meteorological effects. Shrestha's generalized models explained approximately 79\%-85\% of observed variation and found significant effects of temperature, rainfall, humidity, and wind \cite{ref5}. Although multiple particle sizes were analyzed, the results support pollutant-specific meteorological modeling.

Telee et al. found winter and pre-monsoon PM$_{2.5}$ peaks at Ratna Park, negative associations with minimum temperature and precipitation, and a regression model explaining 68\% of variance \cite{ref8}. The authors also noted unmeasured source effects, showing that meteorology modulates pollution but cannot replace information about local and transported emissions.

Using 1,710 daily observations from 2020-2024, Bhatta et al. reported strong seasonality and a random-forest $R^2$ of 0.941 with an RMSE of 12.04 micrograms per cubic meter, substantially exceeding linear regression \cite{ref9}. Rainfall and wind thresholds were associated with lower pollution, supporting nonlinear baselines while cautioning against treating any single threshold as a complete explanation.

At Khumaltar, annual mean PM$_{2.5}$ was 49 micrograms per cubic meter, with winter and pre-monsoon peaks and monsoon reductions. Correlation with NOx indicated an important combustion contribution, while regional transport also affected concentrations \cite{ref10}. These findings support joint modeling of local temporal persistence and cross-location influence.

\subsection{Emission Sources, Biomass Burning, and Regional Transport}

Source-apportionment studies show that Kathmandu PM$_{2.5}$ arises from seasonally varying biomass burning, brick kilns, garbage burning, fossil-fuel combustion, secondary formation, and resuspended material \cite{ref3}, \cite{ref4}. Concentrations and source mixtures differed across urban, suburban, and rural sites, confirming that nearby locations can exhibit materially different pollution behavior.

Wildfire transport creates episodic long-range influence. Kuikel et al. identified 47 days during 2018-2022 with daily PM$_{2.5}$ above 100 micrograms per cubic meter and a maximum of 305; lagged fire counts and backward trajectories supported regional and transboundary transport \cite{ref6}. Static distance graphs cannot fully represent such event-dependent pathways and delays.

Wagle and de Foy separated meteorology, biomass burning, transport, and local-emission effects in hourly PM$_{2.5}$ from 2017-2023. Seasonal temperature explained 36\% of model variance; biomass burning added about 8 micrograms per cubic meter at Phora Durbar, and directional transport clusters changed concentration by roughly +4 or -4 micrograms per cubic meter \cite{ref7}. The decomposition shows why station influence can vary or reverse across meteorological regimes.

Local evidence therefore favors a source-receptor graph: a distant upwind station may be more informative than a nearby crosswind station, and transport lags differ from local accumulation. Edges should consequently vary with wind direction, speed, and time rather than remain permanently undirected.

\section{Evolution of Data-Driven Air-Quality Prediction}

\subsection{Statistical, Machine-Learning, and Ensemble Baselines}

Graph forecasting addresses the fact that distributed urban measurements are dependent but irregularly arranged \cite{ref11}. Conventional models nevertheless remain essential because they are simpler, diagnostically useful, and reveal whether a deep graph architecture adds value beyond nonlinear feature fitting, especially in small or heterogeneous networks.

Singh et al. combined gradient boosting, CatBoost, XGBoost, and LightGBM and interpreted the ensemble with SHAP \cite{ref12}. Its strong single-station performance demonstrates the value of engineered tabular models, but not spatial transfer, pollutant transport, or prediction at unmonitored sites. Tree ensembles should therefore serve as serious nonlinear baselines rather than substitutes for spatial modeling.

Kathmandu studies likewise show that persistence and linear regression are insufficient: nonlinear meteorological models explained 79\%-85\% of variation \cite{ref5}, and random forest outperformed linear regression \cite{ref9}. Evaluation should therefore include persistence, a temporal sequence model, and a tree-based model to isolate gains from temporal memory, nonlinear interactions, and graph message passing.

\subsection{Recurrent and Attention-Based Temporal Modeling}

Recurrent models capture pollutant and meteorological history. GAT-GRU combines graph attention with a GRU and meteorological node features \cite{ref13}, while AAMGCRN embeds graph convolution in LSTM units and fuses it with a separate temporal branch \cite{ref16}. Both illustrate the common separation and subsequent fusion of spatial and temporal learning.

Longer horizons expose the limits of station-only recurrence. A 72-hour GNN-LSTM improved regional $R^2$ from 0.60 to 0.79 by aggregating neighborhood information \cite{ref15}, while a multi-edge graph incorporated seasonal, geographic, domestic, and overseas effects for forecasts up to 72 hours \cite{ref18}. Lead time therefore changes both useful spatial signals and the importance of transported pollution.

Attention-based models improve long-range representation but still lose accuracy with horizon. TKSTAGNet filters irrelevant stations and time steps through Top-K attention \cite{ref19}. A GNN-Transformer model reported MAE of 6.35 and $R^2$ of 0.97 at three hours, compared with MAE of 18.66 and $R^2$ of 0.76 at 24 hours \cite{ref20}, illustrating both attention's value and horizon-dependent degradation.

STGCA combines graph convolution, GRU recurrence, and attention; its RMSE increased across successive forecast steps \cite{ref21}. Because datasets and protocols differ across \cite{ref15}, \cite{ref20}, and \cite{ref21}, their absolute metrics should not be ranked directly. The shared conclusion is that error grows with lead time and must be reported by horizon.

\subsection{Why Spatial Modeling Is Necessary}

Kathmandu wildfire, transport, and source heterogeneity show that a station's own history is not always sufficient \cite{ref3}, \cite{ref4}, \cite{ref6}, \cite{ref7}, \cite{ref10}. Predictive studies similarly report benefits from inter-station modeling through DCRNN, GAT-GRU, AAMGCRN, HighAir, and STAA-GCN \cite{ref13,ref14,ref15,ref16,ref17}, \cite{ref25}, \cite{ref34}.

The central spatial question is how to define an edge. Distance, correlation, attention, contextual similarity, adaptive attributes, hierarchy, and physical transport encode different assumptions about pollutant influence. Graph construction must therefore be treated as a tested model component, not neutral preprocessing.

\section{Spatiotemporal Graph Neural Networks for Air-Quality Forecasting}

\subsection{Graph Representation of Monitoring Networks}

STGNNs combine graph-based spatial encoding with temporal learning \cite{ref11}. Nodes represent stations, cities, or grid cells; node features contain pollutant, meteorological, traffic, land-use, or contextual variables; and weighted edges control message propagation. A temporal module models how these states evolve.

Graphs suit Kathmandu's irregular monitoring network and permit virtual nodes without imposing a false regular grid. However, distance-only edges may miss transport, while fully learned adjacency may capture spurious correlations. Graph design and evaluation should therefore remain explicit experimental choices.

\subsection{Static, Distance-Based, and Correlation-Based Graphs}

Distance-based graphs assume stronger influence among nearby stations and are used directly or alongside fixed correlation structures in several studies \cite{ref14}, \cite{ref18}, \cite{ref21}, \cite{ref23}. They are stable, reproducible, and valuable as baselines, particularly when data are limited.

Their rigidity is the main limitation: fixed undirected edges imply constant reciprocal influence although PM$_{2.5}$ transport is directional and weather-dependent. Correlation can also reflect shared seasonality rather than transport. Monsoon cleaning, winter stagnation, and wildfire arrival can substantially alter the relevant neighborhood \cite{ref5,ref6,ref7,ref8,ref9,ref10}, \cite{ref15}.

\subsection{Attention, Adaptive Adjacency, and Dynamic Edge Attributes}

Graph attention learns unequal neighbor importance. GAT-GRU models city-level dependence \cite{ref13}, ASTGC builds target-specific adjacency from surrounding stations \cite{ref26}, and STAA-GCN fuses proximity, functional, temporal, and complete station relationships \cite{ref34}. These methods reduce reliance on a single manually specified graph.

AAMGCRN learns adjacency from point-of-interest and meteorological data \cite{ref16}, whereas DGN-AEA uses adaptive edge attributes, wind-consistent links, and separate incoming and outgoing transmission \cite{ref17}. Both move from fixed connectivity toward context-dependent graph structure.

Adaptive adjacency is not automatically physically meaningful and may encode dataset-specific co-occurrence. Because reconstructing real network relationships remains difficult \cite{ref17}, learned weights should be compared with transparent distance and wind priors and inspected during known events.

\subsection{Hierarchical and Multi-Scale Spatial Structure}

Pollution transport occurs at multiple spatial scales. HighAir represents this through a city-level graph and station-level graphs, using upper-delivery, lower-updating, and message-passing operations to connect levels. Edge weights are dynamically adjusted by wind direction, and experiments span 10 cities across approximately 61,500 square kilometers \cite{ref25}. The architecture is relevant to Kathmandu because the valley also contains nested spatial structure: neighborhoods and local stations are embedded within a valley-wide circulation system, while regional smoke can enter through external pathways.

SSH-GNN adopts a hierarchical recurrent graph to forecast city-wide air quality at fine granularity. It first approximates a dense pollution distribution from sparse observations and contextual factors, then uses self-supervised objectives to exploit topological and contextual patterns \cite{ref30}. STAA-GCN instead constructs complete station interactions and uses a bi-level sharing strategy to capture common relationship features \cite{ref34}. Both studies show that long-range dependence cannot always be reduced to immediate geographic neighbors, although fully connected or hierarchical graphs require mechanisms to avoid irrelevant or noisy message passing.

\subsection{Temporal Modules, Differential Equations, and Extreme Events}

The selected STGNNs employ several temporal mechanisms: GRU and LSTM recurrence \cite{ref13}, \cite{ref14}, \cite{ref16}, \cite{ref21}, \cite{ref23}, \cite{ref34}; encoder-decoder structures \cite{ref15}, \cite{ref21}, \cite{ref23}, \cite{ref25}; Transformer attention \cite{ref19}, \cite{ref20}; and graph ordinary differential equations \cite{ref22}. The STGODE study argues that shallow graph layers and separate temporal modules may not uniformly represent spatiotemporal dynamics. It incorporates Gaussian diffusion concepts and wind influence into a graph ordinary-differential-equation framework \cite{ref22}. This approach is conceptually attractive for continuous dynamics, although its home-based care-park setting differs substantially from a heterogeneous urban valley.

E-STGCN addresses another underrepresented issue: extreme pollution values. It integrates extreme-value-theory guidance with graph convolution and LSTM sequence modeling, evaluates 37 stations in Delhi, and combines forecasts with conformal prediction intervals \cite{ref33}. This is relevant to Kathmandu because wildfire smoke and winter stagnation produce sharp peaks that average-error objectives can underweight \cite{ref4}, \cite{ref6}, \cite{ref7}. A model that performs well on ordinary hours but systematically misses hazardous peaks would have limited early-warning value.

\section{Wind-Aware and Physics-Guided Modeling}

\subsection{Wind as a Feature Versus Wind as Graph Structure}

The selected literature uses wind in two fundamentally different ways. In feature-aware models, wind speed or direction is appended to the input vector, interpolated over a grid, or fused with other meteorological variables. GAT-GRU includes meteorology as node attributes \cite{ref13}; SwinLSTM-Kriging interpolates wind speed with other regional variables \cite{ref27}; and TKSTAGNet uses a gating module to learn the relative effects of external factors \cite{ref19}. These designs allow wind to influence node states, but they do not necessarily change the direction or strength of information flow between stations.

Structurally wind-aware models modify the graph itself. The DCRNN study combines geographic distance with dominant wind direction and reports that the directed graph outperforms the undirected alternative for 24-hour PM$_{2.5}$ and ozone forecasts, whereas the undirected graph can be more effective at the immediate one-hour horizon \cite{ref14}. This horizon-dependent result is physically plausible: local persistence and nearby concentration dominate very short horizons, while directional transport becomes more influential as lead time increases.

The 72-hour GNN-LSTM analysis provides complementary evidence. The contribution of distant sites associated with regional transport increased from approximately 0\% to 38\% across the 72-hour horizon, and upwind neighboring stations were particularly important during polluted episodes \cite{ref15}. HighAir dynamically adjusts hierarchical edge weights according to wind direction \cite{ref25}, while DGN-AEA retains wind-consistent edges and models incoming and outgoing transmission separately \cite{ref17}. The spatial-diffusion encoder-decoder and STGODE studies similarly integrate diffusion or wind fields into graph operations \cite{ref22}, \cite{ref23}. Collectively, these papers support a dynamic directed graph rather than a graph in which wind is only one additional scalar feature.

For Kathmandu, a practical wind-aware graph should encode both direction and magnitude. Direction determines whether station j is plausibly upwind of station i, while speed affects transport strength and potential time lag. A distance-decay term can preserve local influence, but its weight should be modulated by wind alignment. Because valley topography may channel or obstruct flow, the resulting graph should be treated as a testable approximation rather than a complete atmospheric model. The project should compare at least three graph variants: distance-only, learned/adaptive, and wind-aware dynamic adjacency.

\subsection{Physics-Guided Neural Networks}

Physics-guided models attempt to constrain learning with transport processes rather than relying entirely on statistical association. AirPhyNet represents diffusion and advection through differential-equation networks and combines them with a graph architecture. Across two benchmark datasets, it evaluated 24-, 48-, and 72-hour horizons, sparse data, and sudden-change conditions, reporting error reductions of up to 10\%. Its case study also argues that learned behavior corresponds to underlying particle movement \cite{ref24}. The methodological value lies less in a single performance number than in the use of physical structure to improve long-horizon and sparse-data robustness.

PGITS applies physics guidance to inductive spatiotemporal kriging. A dynamic graph-generation module incorporates advection and diffusion into the adjacency matrix, and an increment-training strategy attempts to reduce the mismatch between simulated virtual nodes and actual unobserved nodes \cite{ref31}. SPIN extends this direction with parallel advection and diffusion graph kernels, dynamic node masking, and a satellite AOD spatial-gradient constraint. Rather than using incomplete AOD directly as a feature, SPIN uses its spatial gradient in the loss and reports an MAE of 9.52 micrograms per cubic meter for continuous inference in the Beijing-Tianjin-Hebei region \cite{ref32}.

Physics guidance offers three potential advantages for the present project: it can regularize edge learning when station data are sparse; it can improve interpretability by connecting edge behavior to transport; and it can support extrapolation to unmonitored locations. Its risks are equally important. Simplified advection-diffusion assumptions may omit complex terrain circulation, vertical mixing, chemical transformation, and emission variability. Accordingly, a lightweight wind-aware inductive bias is more defensible for an undergraduate project than claiming to reproduce full atmospheric physics. The value of the bias should be established through ablation experiments, event-based inspection, and comparison with simpler graphs.

\subsection{Design Requirements Derived from Wind and Transport Studies}

\begin{itemize}
    \item Represent the graph as directed and time-varying, while retaining a reproducible distance-only baseline.
    \item Encode wind direction in a circular or vector form so that directions near 0 and 360 degrees are treated as adjacent.
    \item Allow the importance of distant upwind stations to increase with forecast horizon and transport conditions.
    \item Inspect edge weights during winter stagnation, monsoon cleaning, and wildfire episodes rather than reporting only average test error.
    \item Separate predictive association from causal interpretation; an attention weight is evidence of model reliance, not proof of physical transport.
    \item Evaluate whether the wind-aware graph improves hazardous-episode forecasts, not only ordinary-hour averages.
\end{itemize}

\section{Forecasting and Estimation at Unmonitored Locations}

\subsection{Regional Interpolation and Fine-Grained Surface Prediction}

Several selected studies convert sparse station observations into regional surfaces before or during prediction. SwinLSTM-Kriging interpolates meteorological and pollutant indicators from 29 stations into two-dimensional regional fields and then predicts PM$_{2.5}$ for 1-, 6-, and 24-hour horizons. The proposed SwinLSTM improved average $R^2$ by approximately 5\% over ConvLSTM \cite{ref27}. The method demonstrates that interpolation can create a regular spatial representation for regional forecasting, but prediction quality remains dependent on the assumptions and smoothing behavior of Kriging.

ASTGC directly addresses fine-grained AQI estimation when monitoring stations are sparse. It separates dynamic temporal features from static spatial context and uses attention to construct a target-specific adjacency matrix from surrounding stations \cite{ref26}. STGCA uses satellite remote-sensing information with graph convolution and seq2seq prediction to generate high-resolution PM$_{2.5}$ surfaces \cite{ref21}. These approaches show that regional mapping can be integrated with forecasting, although AQI prediction and concentration prediction should remain conceptually distinct. For the present project, PM$_{2.5}$ should be estimated first and converted to AQI afterward so that model error is not hidden by categorical or piecewise AQI transformations.

\subsection{Hidden Stations, Sparse Stations, and Inductive Generalization}

AQ-Net is designed for reanalysis at both observed and unobserved stations. It combines LSTM and multi-head attention for temporal regression with a neural k-nearest-neighbor module for feature-based interpolation. Its evaluation distinguishes visible stations that provide historical measurements from hidden stations for which only coordinates are supplied \cite{ref28}. The sparse-station hybrid model by Hu et al. instead extracts multi-scale meteorological lags, selects multi-source features, divides related stations by spatial hierarchy, and reports advantages for sudden changes and long-term prediction in sparse networks \cite{ref29}. Both papers emphasize that models optimized for dense station networks may not transfer reliably to sparse settings.

SSH-GNN addresses future forecasting over unmonitored regions rather than only reconstructing current conditions. It first approximates city-wide air quality using historical readings and contextual factors such as weather and traffic, then applies a hierarchical recurrent GNN with self-supervised topological and contextual objectives \cite{ref30}. PGITS and SPIN address inductive kriging more directly through virtual-node or masked-node training \cite{ref31}, \cite{ref32}. Their central idea is that a model must learn a transferable relationship between observed nodes and locations that do not supply pollutant history.

\subsection{Two Different Unmonitored-Site Tests}

The literature reveals that the term unmonitored can describe different experimental conditions. A hidden real station may have known coordinates, meteorological history, static context, or even pollutant information available during some stages of training. A completely new map location may have coordinates and meteorology but no historical PM$_{2.5}$ record and may not have appeared as a node during model fitting. These tasks have different difficulty levels and different leakage risks.

The proposed project should therefore report two separate experiments. In the hidden-station experiment, one or more real stations should be withheld spatially, and their PM$_{2.5}$ values must not enter feature construction, graph learning, normalization statistics, or target interpolation during the evaluation period. Leave-one-station-out or grouped spatial cross-validation is preferable to random row splitting because random splitting preserves station identity in both training and test sets. In the new-location experiment, a virtual node should be inserted using only information that would genuinely be available at a sensor-free location, such as coordinates, meteorology, and static context. Prediction uncertainty should be expected to increase with distance from observed stations and under novel meteorological conditions.

The selected studies operationalize an unseen location differently \cite{ref26,ref27,ref28,ref29,ref30,ref31,ref32}. The project must therefore report hidden-station and new-location results separately to avoid presenting interpolation as unrestricted spatial generalization.

\section{Explainability, Extreme Events, and Operational Reliability}

\subsection{Feature Attribution and Model Transparency}

The literature provides three complementary forms of explanation. First, interpretable statistical models quantify smooth nonlinear relationships between PM$_{2.5}$ and drivers. The Kathmandu GAM study separates seasonal temperature, biomass burning, transport direction, diurnal behavior, and boundary-layer influence \cite{ref7}, while the local generalized-linear analysis quantifies meteorological associations and lags \cite{ref5}. These models provide valuable domain checks for a more complex neural system.

Second, SHAP-based analysis attributes a prediction to individual input features. Singh et al. use SHAP to interpret a weighted ensemble and identify the contribution of pollutants and meteorological variables \cite{ref12}. For the proposed project, SHAP can be applied globally to summarize the importance of historical PM$_{2.5}$, wind, temperature, humidity, precipitation, and other inputs, and locally to explain a specific high-pollution forecast. However, attribution values describe the model's use of features; they do not prove that changing a feature would cause the predicted physical outcome.

Third, graph attention and edge weights can identify which neighboring stations contribute to a target prediction. GAT-GRU, ASTGC, TKSTAGNet, HighAir, and STAA-GCN all include mechanisms that emphasize selected nodes or relationships \cite{ref13}, \cite{ref19}, \cite{ref25}, \cite{ref26}, \cite{ref34}. These weights are useful for checking whether the model focuses on plausible upwind or functionally related stations. They should not be presented as complete explanations because attention can be distributed across layers, heads, and time steps and may not correspond directly to feature sensitivity.

A stronger XAI design combines these views. SHAP should describe input-feature contribution; graph analysis should describe station-to-station influence; and meteorological context should assess whether the explanation is physically plausible. For example, a forecast attributed to a distant station should be examined against wind alignment and travel time. Explanations should also be stratified by forecast horizon, because the 72-hour study \cite{ref15} and wind-directed DCRNN \cite{ref14} show that spatial influence changes as the lead time increases.

\subsection{Extreme Pollution and Uncertainty}

Average MAE or RMSE can obscure the most important public-health failures. The wildfire study documents PM$_{2.5}$ peaks up to 305 micrograms per cubic meter \cite{ref6}, and winter measurements show sustained high concentrations across urban, suburban, and rural sites \cite{ref4}. E-STGCN explicitly incorporates extreme-value behavior and supplements point forecasts with conformal prediction intervals \cite{ref33}. Although implementing full extreme-value modeling may exceed the initial project scope, the project should at minimum report performance for high-pollution subsets, seasonal peaks, and rapid concentration changes.

Operational reliability also requires calibrated communication. AQI should be derived from predicted PM$_{2.5}$ using the selected reporting standard, but concentration-level metrics should remain primary. A forecast that crosses an AQI category boundary can have a large communication effect even when the numerical concentration error is modest. Conversely, a model may achieve good AQI-category accuracy while retaining substantial concentration bias. Reporting both outputs avoids this ambiguity.

\section{Comparative Synthesis and Research Gaps}

\subsection{What the Literature Establishes}

The selected evidence establishes that air-quality forecasting is inherently spatiotemporal. Kathmandu measurements exhibit seasonality, diurnal structure, nonlinear meteorological response, local-source effects, and regional transport \cite{ref1,ref2,ref3,ref4,ref5,ref6,ref7,ref8,ref9,ref10}. Graph-based studies show that incorporating neighboring stations generally improves prediction relative to station-independent temporal models, particularly at longer horizons \cite{ref13,ref14,ref15,ref16,ref17,ref18,ref19,ref20,ref21,ref22,ref23,ref24,ref25}, \cite{ref34}. Attention and adaptive adjacency improve flexibility, while hierarchical models represent interactions at multiple spatial scales \cite{ref16}, \cite{ref17}, \cite{ref19}, \cite{ref25}, \cite{ref30}, \cite{ref34}. Physics-guided models further suggest that advection and diffusion priors can improve sparse-data and long-horizon behavior \cite{ref24}, \cite{ref31}, \cite{ref32}.

The literature also establishes that unmonitored-site inference is feasible but evaluation-sensitive. Kriging, neural kNN, target-specific attention, dense city-wide approximation, virtual-node training, and masked inductive learning all provide mechanisms for spatial estimation \cite{ref26,ref27,ref28,ref29,ref30,ref31,ref32}. However, the meaning of unmonitored varies across studies. Without a strict spatial holdout and a clear statement of what target-node information is available, reported generalization can reflect interpolation within a known network rather than deployment at a genuinely new location.

\subsection{Unresolved Gaps Relevant to the Project}

First, there is a geographic and environmental transfer gap. The Kathmandu papers provide detailed observational and source evidence but do not present a mature wind-aware STGNN evaluated across the valley. Conversely, most advanced graph models are validated in Chinese, Indian, or Korean networks with different station densities, terrain, emissions, and meteorological regimes \cite{ref13,ref14,ref15,ref16,ref17,ref18,ref19,ref20,ref21,ref22,ref23,ref24,ref25,ref26,ref27,ref28,ref29,ref30,ref31,ref32,ref33,ref34}. The proposed project therefore requires local validation rather than assuming that an architecture successful elsewhere will transfer directly.

Second, wind is frequently included as a node feature or external variable but less consistently used to control directed, time-varying connectivity. The strongest structural evidence comes from DCRNN, the 72-hour GNN-LSTM, DGN-AEA, HighAir, STGODE, the spatial-diffusion encoder-decoder, AirPhyNet, PGITS, and SPIN \cite{ref14}, \cite{ref15}, \cite{ref17}, \cite{ref22,ref23,ref24,ref25}, \cite{ref31}, \cite{ref32}. Even among these, implementations differ substantially, and few are evaluated in a topographically confined valley. A direct comparison between distance-only, learned, and wind-aware graphs remains necessary.

Third, forecasting at monitored stations and estimation at unmonitored locations are often addressed separately. Some methods forecast future values for known nodes, while others reconstruct current or historical values at hidden locations. SSH-GNN, PGITS, and SPIN move toward future or inductive inference \cite{ref30,ref31,ref32}, but the selected literature still leaves room for a unified model evaluated under both hidden-station and fully new-location conditions.

Fourth, explainability is uneven. SHAP is used in the ensemble study \cite{ref12}, physical case analysis is emphasized by AirPhyNet \cite{ref24}, and graph attention is present in several models. Nevertheless, few studies jointly explain feature effects, temporal dependence, and spatial station influence. The proposed project can address this by presenting feature attribution and wind-conditioned graph influence as separate but complementary evidence.

Fifth, experimental comparability remains limited. Reported metrics span AQI and pollutant concentration, one-step and 72-hour horizons, single-station and regional tasks, random and temporal splits, and different normalization procedures. A lower RMSE in one paper does not imply a universally superior model. The project's evaluation must use identical data splits, horizons, baselines, and metrics for every compared architecture. Ablation studies should isolate the value of meteorology, dynamic wind edges, graph attention, and virtual-node mechanisms.

Finally, several highly relevant selected papers are preprints or early-access versions, including the 72-hour GNN-LSTM, AQ-Net, the spatial-diffusion encoder-decoder, HighAir, PGITS, SPIN, and E-STGCN \cite{ref15}, \cite{ref23}, \cite{ref25}, \cite{ref28}, \cite{ref31,ref32,ref33}. Their methods are valuable and current, but their findings should be treated as promising rather than definitive. The literature review therefore uses them to motivate design alternatives while relying on local observational evidence and peer-reviewed graph studies for the strongest claims.

\section{Positioning of the Proposed Project}

The proposed project is positioned as a locally validated integration of ideas that are individually supported by the selected literature. Its principal target is hourly PM$_{2.5}$ concentration across Kathmandu Valley. AQI is produced only after concentration prediction, preserving a continuous scientific target for model training and error analysis. A 24-hour historical input window is used to generate 1-, 6-, 12-, and 24-hour forecasts, with six hours treated as the main operational horizon. This design reflects the observed deterioration of long-horizon accuracy \cite{ref15}, \cite{ref18}, \cite{ref20}, \cite{ref24} while preserving both immediate and day-ahead use cases.

Spatially, real monitoring stations are represented as graph nodes. The base graph can use geographic distance, while the proposed wind-aware graph modifies edge direction and strength according to time-aligned wind conditions. This choice is grounded in Kathmandu transport evidence \cite{ref6}, \cite{ref7}, \cite{ref10} and graph-model findings showing the value of directed wind relationships \cite{ref14}, \cite{ref15}, \cite{ref17}, \cite{ref25}. An adaptive or attention-based component can refine the prior graph, but the physical wind term should remain separately inspectable so that improved performance can be attributed to a defensible mechanism.

The project also extends beyond conventional station forecasting. For a hidden real station, PM$_{2.5}$ values are withheld to test whether the model can infer a known location from neighboring stations and meteorology. For a completely new map location, the model receives no PM$_{2.5}$ history and must rely on spatial context, meteorology, and learned transfer from observed nodes. This two-part protocol draws on AQ-Net, SSH-GNN, PGITS, and SPIN \cite{ref28}, \cite{ref30,ref31,ref32} but avoids treating the two tasks as equivalent.

Explainability is integrated at two levels. SHAP is used to quantify the contribution of pollutant history and meteorological features, following the interpretable ensemble direction in \cite{ref12}. Graph-attention or edge-weight analysis is used to show which stations influence each target and whether those stations are meteorologically plausible contributors. The local interpretable studies \cite{ref5}, \cite{ref7} provide an external reasonableness check: feature effects should be compared with known seasonal, wind, precipitation, fire, and transport behavior rather than interpreted in isolation.

The expected contribution is therefore not a claim that graph neural networks, wind, virtual nodes, or SHAP are individually unprecedented. The defensible contribution is their unified and carefully evaluated application to a sparse, topographically constrained, and transport-sensitive Kathmandu Valley setting. The project's quality will depend less on architectural complexity than on leakage-free data preparation, transparent graph construction, strong baselines, horizon-specific evaluation, spatial generalization tests, and critical interpretation of failure cases.

\section{Chapter Summary}

This chapter overviews 34 selected studies on Kathmandu PM$_{2.5}$ behavior, spatiotemporal prediction, graph construction, wind and physical transport, unmonitored-site inference, and explainability. Local evidence demonstrates pronounced seasonal and spatial variability, nonlinear meteorological effects, heterogeneous sources, and regional transport. These conditions make station-independent models and fixed undirected graphs insufficient as the sole problem representation. The methodological literature shows that STGNNs represent irregular monitoring networks through spatial message passing combined with recurrent, attention-based, Transformer, or differential-equation temporal modules. Directed wind graphs, adaptive edges, hierarchical structure, and physics-guided transport are especially relevant to longer horizons. Inductive studies also show that unmonitored-site estimation is feasible, provided hidden-station reconstruction is distinguished from prediction at a new location without PM$_{2.5}$ history. These gaps support a wind-aware, multi-horizon STGNN for monitored and unmonitored Kathmandu locations, with separate feature and spatial explanations.
